Visualizing probability distributions across bivariate cyclic temporal granularities

### 2020-08-05 main.Rmd
smart meter data minor changes proof-read

For data sets of the form ($C_i$, $C_j$, $v$) $\forall i, j \in N_C$, if there are levels of $C_i$ (or $C_j$) not spanned by levels of $C_j$ (or $C_i$), empty sets are formed leading to potential ineffective graphs. The conjecture is that the synergy of these cyclic granularities are thus playing a role while deciding if the resulting plot would be a good candidate for exploratory analysis. Harmonies are pairs of granularities that do not contain empty combinations and thus aid exploratory data analysis. Plotting Clashes should be avoided since they contain empty combinations. For illustration, \autoref{fig:allFig} (a) shows the distribution of half-hourly electricity consumption through letter value plot across months of the year faceted by quarters of the year. This plot does not work because every quarter does not correspond to all months of the year, for example, the first quarter of the year would never correspond to the month December of an year.

<!--In Section \ref{sec:synergy}, we discussed how pairs of granularities can have empty combinations either due to structure of calendar, event location or duration or due to the way they are built. In this section, we will see how these empty combinations affect the visualization when a dependent variable is plotted against these granularities.-->

<!---\autoref{fig:allFig} (c) shows box plot across days of the year by the 1st, 15th, 29th and 31st days of the month. \autoref{fig:allFig} (d) showing violin plot across days of the month faceted by week of the month. \autoref{fig:allFig} (e), variations across week of the year conditional on week of the month can be observed through a ridge plot and \autoref{fig:allFig}(b) shows decile plots across day of the year and month of the year.--->

<!-- Clearly, in \autoref{fig:allFig}, we observe that  -->

<!--In (b) we can not compare the distributions across different facets because every day-of-year cannot occur for every month. For example, 21st day of the year can never correspond to the month March or December of any year.-->
<!---In \autoref{fig:allFig} (c), there will be no observations for some combinations day-of-month and day-of-year. In particular, the 1st day of the month can never correspond to 2nd, 3rd or 4th day of the year. On the contrary, for \autoref{fig:allFig}--->  <!---Thus the graphs that don't work are those where many of the combination sets are empty.--->

<!---as they hinder the exploratory process by having missing combinations of time granularities in each panel.--->

@@

Visualizing distributions can be misleading if statistical transformations are performed on rarely occurring categories (Section \ref{sec:summarize-measured}) or unevenly distributed events. For Kernel density based plots $nobs>30$ is suggested, for summary statistics based plots like boxplots $nobs = 10$ is acceptable and for quantile plots $nobs$ should at least be equal to the number of quantiles plotted.

<!-- If we require at least $k$ observations per combination to create a meaningful summarization, then only $T \ge lmk$ will be acceptable. The value of $k$ will depend on the statistical transformation we are producing. For computing deciles, $k=10$ may be acceptable, but for density estimates, $k \ge 30$ is suggested.  -->

@@

Even when there are no rarely occurring events, number of observations might vary hugely within or across each facet. This might happen due to missing observations in the data or uneven locations of events in time domain. In such cases, the statistical transformations based on density should be used with caution as sample sizes would directly affect both the variance and consequently the confidence interval of the estimators. <!-- (see \autoref{eq:MSE})  shows that the variance of both $\hat{f(x)}$ <!--and $\hat{q(u)}$is a function of the sample size $n$.-->

@@

The tsibble object `smart_meter10` from R package `gravitas` [@R-gravitas] consisting of `reading_datetime`, `customer_id` and `general_supply_kwh` denoting the index, key and measured variable of the tsibble is used to facilitate the systematic exploration. While trying to explore the energy behavior of these customers systematically across cyclic time granularities, the first thing to consider is which cyclic time granularities we can look at exhaustively. Let us consider conventional time deconstructions for a Gregorian calendar (second, minute, half-hour, hour, day, week, month, year). Since the interval of this tsibble is 30 minutes, the temporal granularities may range from half-hour to year. Considering $6$ linear granularities half-hour, hour, day, week, month and year in the hierarchy table, $N_C = (6*5/2) = 15$. <!--(via Section \ref{sec:data-structure}) circular or quasi-circular granularities that could be formed relating two linear granularities at a time.--> If $N_C$ seem too large, the smallest and largest linear granularities could be considered to be removed from the hierarchy table. We remove half-year and year to have $N_C = (4*3/2) = 6$ and obtain cyclic granularities namely "hour_day", "hour_week", "hour_month", "day_week", "day_month" and "week_month", read as "hour of the day", etc. Further, we add cyclic granularity day-type( "wknd_wday") to capture weekend and weekday behavior. Now that we have a list of cyclic granularities to look at, we should be able to compute the multiple-order-up granularities using Section \ref{sec:cyclic-calendar}. <!-- The lubridate package [@Grolemund2011-vm] provides tools to compute single-order-up granularities, \ref{sec:circular-gran-def} (for hour_day, hour_week and day_week) and \ref{sec:quasi-circular-gran-def} (for hour_month, day_month, week_month).  
with the index set mapping half-hourly time periods to a set of positive integers.-->

@@

From the search list, $N_C = 7$ cyclic granularities are chosen for which we would like to derive insights of energy behavior. Recalling the data structure <$C_i$, $C_j$, `general_supply_kwh`> for exploration $\forall i, j \in \{1, 2, \ldots, 7\}$, each of these $7$ cyclic granularities can either be mapped to x-axis or to facet. Choosing $2$ of the possible $7$ granularities, which is equivalent to having $^{7}P_2 = 42$ candidates for visualization. Fortunately, harmonies can be identified among those $42$ possibilities to narrow the search. \autoref{tab:harmony-tab} shows $16$ harmony pairs after removing clashes and any cyclic granularities with levels more than $31$, as effective  exploration becomes difficult with many levels (Section \ref{sec:levels}). The MMPD is also shown along with indicator (*) only when variation of measured variable across the harmony pair significant. Starting from $42$ possible pairs of cyclic granularities to visualize, we are finally left with only $6$, which is a very sizable number of displays for exploration.

Few harmony pairs are displayed in \autoref{fig:bothcust} to illustrate the siginificance of MMPD, threshold and the impact of different distribution plots and reverse mapping.
\noindent For each of \autoref{fig:bothcust} (b) and (c), $C_i$ is the circular granularity day-type (weekday/weekend) and $C_j$ is hour of the day. The geometry used for displaying the distribution is chosen as area-quantiles and violins in \autoref{fig:bothcust} (b and c respectively). \autoref{fig:bothcust} (a) displays reverse mapping of $C_i$ and $C_j$ with $C_i$ denoting hour of the day and $C_j$ denoting day-type with distribution geometrically displayed as boxplots.

@@

  harmonies <- sm %>%
  harmony(ugran = "month",
          filter_in = "wknd_wday", 
          filter_out = c("hhour", "fortnight"))
set.seed(12345)

global_harmony <- sm %>%
  global_threshold(harmony_tbl = harmonies,
                   response = "general_supply_kwh", nsamp = 5) %>%
  rename(`facet variable` = facet_variable,
         `x-axis variable` = x_variable,
         `facet levels` = facet_levels,
         `x-axis levels` = x_levels) %>% 
  mutate(MMPD = if_else(gt_MMPD == TRUE,
                        paste0(MMPD, "*"), paste0(MMPD))) %>%
  select(`facet variable`,`x-axis variable`,
         `facet levels`,`x-axis levels` ,MMPD)

#   write_rds(smart_harmony,"cache/smart_harmony_ak.rds")
# }
#smart_harmony <- read_rds("cache/smart_harmony.rds")

  knitr::kable(global_harmony,
               format = "latex",
               booktabs = TRUE,
                caption = "Harmonies with a pair of cyclic granularity one placed on facet and the other on x-axis. Out of 42 possible combinations of cyclic granularities, only 16 are harmony pairs. The harmony pairs are ranked basis highest to lowest MMPD and only the first 6 show significant variation in the measured variable.") %>%
    #row_spec(0, bold = TRUE) %>%
    kable_styling()
```

@@

In \autoref{fig:bothcust} (b), <!--the distribution of energy consumption is plotted across the harmony pair (weekday/weekend, hour of the day) through an area quantile plot-->the black line is the median, whereas the purple band covers 25th to 75th percentile, the orange band covers 10th to 90th percentile and the green band covers 1st to 99th percentile. The first facet represents the weekday behavior while the second one displays the weekend behavior and energy consumption across each hours of the day is shown inside each facet. The energy consumption is extremely (positive- or right-) skewed with the 1st, 10th and 25th percentile lying relatively close whereas 75th, 90th and 99th lying further away from each other. This is common across both weekdays and weekends. For the first few hours on weekdays, median energy consumption starts and continues to be higher for longer as compared to weekends.

Consider looking at violin plots instead of quantile plots to look at the same data in \autoref{fig:bothcust}(c). There is additional information that we can derive looking at the distribution. There is bimodality in the early hours of the day, implying both low and high energy consumption is probable in the early hours of the day both for weekdays and weekends. <!--Also the hours from 7 to 13 look most volatile.--> If we visualize the same data with reverse mapping of the cyclic granularities, then the natural tendency would be to compare weekend and weekday behavior within each hour and not across hours. For example in \autoref{fig:bothcust}(a), it can be seen that median energy consumption for the early morning hours is extremely high for weekdays compared to weekends. Also, outliers are more prominent in the latter part of the day. All of these indicate that looking at different distribution geometry or changing the mapping might shed lights on different aspect of the energy behavior for the same sample population.

@@

If the data for all keys are visualized together, it might lead to Simpson's paradox, which occurs when one observation shows a particular behavior, but this behavior paradoxically becomes obscured by aggregation. For example in a particular neighborhood one household may have the least daily power consumption for a full week, yet still not be the household with the minimum weekly power consumption. This is an intuitive possibility, because heterogeneous `customer_id`'s with very different occupation or demographics will tend to have very different energy behavior and combining them together will somehow weaken any typical or extreme behavior. A strategy for analyzing multiple keys together could be to first group them basis time series or demographic features and then look at their energy behavior. This is beyond the scope of the current work.

@@

This case study shows systematic exploration of energy behavior for a household to gain exhaustive insights on periodic behavior of the households. 
<!--First, it helps us to find the list of cyclic granularities to look at, then shrinks the number of possible visualizations by identifying harmonies, visualize a harmony pair and shows the effect of different distribution plots or reverse mapping.-->

>>>>

For data sets of the form ($C_i$, $C_j$, $v$) $\forall i, j \in N_C$, if there are levels of $C_i$ (or $C_j$) not spanned by levels of $C_j$ (or $C_i$), empty sets are formed leading to potential ineffective graphs. The conjecture is that the synergy of these cyclic granularities are thus playing a role while deciding if the resulting plot would be a good candidate for exploratory analysis. Harmonies are pairs of granularities that do not contain empty combinations and could possibly be useful for exploring patterns. However, plotting clashes should be avoided since they might contain combinations which are empty or have too few observations. For illustration, \autoref{fig:allFig} (a) shows the distribution of half-hourly electricity consumption through letter value plot across months of the year faceted by quarters of the year. This plot does not work because every quarter does not correspond to all months of the year, for example, the first quarter of the year would never correspond to the month December of an year.

@@

Visualizing distributions can be misleading if statistical transformations are performed on rarely occurring categories (Section \ref{sec:summarize-measured}). Even when there are no rarely occurring events, number of observations might vary hugely within or across each facet. This might happen due to missing observations in the data or uneven locations of events in time domain. In such cases, the statistical transformations should be used with caution as sample sizes would directly affect both the variance and consequently the confidence interval of the estimators.

<!-- For Kernel density based plots $nobs>30$ is suggested, for summary statistics based plots like boxplots $nobs = 10$ is acceptable and for quantile plots $nobs$ should at least be equal to the number of quantiles plotted. If we require at least $k$ observations per combination to create a meaningful summarization, then only $T \ge lmk$ will be acceptable. The value of $k$ will depend on the statistical transformation we are producing. For computing deciles, $k=10$ may be acceptable, but for density estimates, $k \ge 30$ is suggested.  -->

@@

<!-- (see \autoref{eq:MSE})  shows that the variance of both $\hat{f(x)}$ <!--and $\hat{q(u)}$is a function of the sample size $n$.-->

@@

The tsibble object `smart_meter10` from R package `gravitas` [@R-gravitas] consisting of `reading_datetime`, `customer_id` and `general_supply_kwh` denoting the index, key and measured variable of the tsibble is used to facilitate the systematic exploration. While trying to explore the energy behavior of these customers systematically across cyclic time granularities, the first thing to consider is which cyclic time granularities we can look at exhaustively. Let us consider conventional time deconstructions for a Gregorian calendar (second, minute, half-hour, hour, day, week, month, year). Since the interval of this tsibble is 30 minutes, the temporal granularities may range from half-hour to year. Considering $6$ linear granularities half-hour, hour, day, week, month and year in the hierarchy table, $N_C = (6*5/2) = 15$. <!--(via Section \ref{sec:data-structure}) circular or quasi-circular granularities that could be formed relating two linear granularities at a time.--> If $N_C$ seem too large, the smallest and largest linear granularities could be considered to be removed for analysis. Granularities half-year and year are removed to have $N_C = (4*3/2) = 6$ and obtain cyclic granularities namely "hour_day", "hour_week", "hour_month", "day_week", "day_month" and "week_month", read as "hour of the day", etc. Further, we add cyclic granularity day-type( "wknd_wday") to capture weekend and weekday behavior. Now that we have a list of cyclic granularities to look at, we should be able to compute them using Section \ref{sec:cyclic-calendar}. <!-- The lubridate package [@Grolemund2011-vm] provides tools to compute single-order-up granularities, \ref{sec:circular-gran-def} (for hour_day, hour_week and day_week) and \ref{sec:quasi-circular-gran-def} (for hour_month, day_month, week_month).  
with the index set mapping half-hourly time periods to a set of positive integers.-->

@@

From the search list, $N_C = 7$ cyclic granularities are chosen for which we would like to derive insights of energy behavior. Recalling the data structure <$C_i$, $C_j$, `general_supply_kwh`> for exploration $\forall i, j \in \{1, 2, \ldots, 7\}$, each of these $7$ cyclic granularities can either be mapped to x-axis or to facet. Choosing $2$ of the possible $7$ granularities, which is equivalent to having $^{7}P_2 = 42$ candidates for visualization. Harmonies can be identified among those $42$ possibilities to narrow the search. \autoref{tab:harmony-tab} shows $16$ harmony pairs after removing clashes and any cyclic granularities with levels more than $31$, as effective  exploration becomes difficult with many levels (Section \ref{sec:levels}).

<!--The MMPD is also shown along with indicator (*) only when variation of measured variable across the harmony pair significant. Starting from $42$ possible pairs of cyclic granularities to visualize, we are finally left with only $6$, which is a very sizable number of displays for exploration.-->

\noindent Few harmony pairs are displayed in \autoref{fig:bothcust} to illustrate the <!--siginificance of MMPD, threshold and the -->impact of different distribution plots and reverse mapping. For each of \autoref{fig:bothcust} (b) and (c), $C_i$ is the circular granularity day-type (weekday/weekend) and $C_j$ is hour of the day. The geometry used for displaying the distribution is chosen as area-quantiles and violins in \autoref{fig:bothcust} (b and c respectively). \autoref{fig:bothcust} (a) displays reverse mapping of $C_i$ and $C_j$ with $C_i$ denoting hour of the day and $C_j$ denoting day-type with distribution geometrically displayed as boxplots. 
In \autoref{fig:bothcust} (b), <!--the distribution of energy consumption is plotted across the harmony pair (weekday/weekend, hour of the day) through an area quantile plot-->the black line is the median, whereas the purple band covers 25th to 75th percentile, the orange band covers 10th to 90th percentile and the green band covers 1st to 99th percentile. The first facet represents the weekday behavior while the second one displays the weekend behavior and energy consumption across each hours of the day is shown inside each facet. The energy consumption is extremely (positive- or right-) skewed with the 1st, 10th and 25th percentile lying relatively close whereas 75th, 90th and 99th lying further away from each other. This is common across both weekdays and weekends. For the first few hours on weekdays, median energy consumption starts and continues to be higher for longer as compared to weekends.

\noindent Consider looking at violin plots instead of quantile plots to look at the same data in \autoref{fig:bothcust}(c) for some more insights. There is bimodality in the early hours of the day, implying both low and high energy consumption is probable in the early hours of the day both for weekdays and weekends. <!--Also the hours from 7 to 13 look most volatile.--> If we visualize the same data with reverse mapping of the cyclic granularities, then the natural tendency would be to compare weekend and weekday behavior within each hour and not across hours. For example in \autoref{fig:bothcust}(a), it can be seen that median energy consumption for the early morning hours is higher for weekdays than weekends. Also, outliers are more prominent in the latter hours of the day. All of these indicate that looking at different distribution geometry or changing the mapping might shed lights on different aspect of the energy behavior for the same sample population.

@@

  harmonies <- sm %>%
  harmony(ugran = "month",
          filter_in = "wknd_wday", 
          filter_out = c("hhour", "fortnight")) %>%
    rename(`facet variable` = facet_variable,
         `x-axis variable` = x_variable,
         `facet levels` = facet_levels,
         `x-axis levels` = x_levels)

# set.seed(12345)
# 
# global_harmony <- sm %>%
#   global_threshold(harmony_tbl = harmonies,
#                    response = "general_supply_kwh", nsamp = 5) %>%
#   rename(`facet variable` = facet_variable,
#          `x-axis variable` = x_variable,
#          `facet levels` = facet_levels,
#          `x-axis levels` = x_levels) %>% 
#   mutate(MMPD = if_else(gt_MMPD == TRUE,
#                         paste0(MMPD, "*"), paste0(MMPD))) %>%
#   select(`facet variable`,`x-axis variable`,
#          `facet levels`,`x-axis levels`)

  knitr::kable(harmonies,
               format = "latex",
               booktabs = TRUE,
                caption = "Harmonies with a pair of cyclic granularity one placed on facet and the other on x-axis. Out of 42 possible combinations of cyclic granularities, only 16 are harmony pairs.") %>%
    #row_spec(0, bold = TRUE) %>%
    kable_styling()
```

@@

@@

If the data for all keys are visualized together, it might lead to Simpson's paradox, which occurs when one observation shows a particular behavior, but this behavior paradoxically becomes obscured by aggregation. For example in a particular neighborhood one household may have the least daily power consumption for a full week, yet still not be the household with the minimum weekly power consumption. This is an intuitive possibility, because heterogeneous `customer_id`'s with very different occupation or demographics will tend to have very different energy behavior and combining them together will somehow weaken any typical or extreme behavior. A strategy for analyzing periodicities of multiple keys together could be to group the keys basis their energy behavior across different cyclic granularities. This is beyond the scope of the current work.

@@

<!--This case study shows systematic exploration of energy behavior for a household to gain exhaustive insights on periodic behavior of the households. -->

<!--First, it helps us to find the list of cyclic granularities to look at, then shrinks the number of possible visualizations by identifying harmonies, visualize a harmony pair and shows the effect of different distribution plots or reverse mapping.-->
